---
title: "CLN3 Lysosomal Interactome — Proteomics Analysis"
subtitle: "Dataset: ProteomeXchange PXD031582"
author: "Santiago López-Begines"
date: "`r Sys.Date()`"
output:
  rmdformats::material:
    number_sections: TRUE
    code_folding: show
    highlight: tango
    fig_width: 10
    fig_height: 8
    self_contained: TRUE
    cards: TRUE
editor_options:
  markdown:
    wrap: sentence
---

# Importing Data

This analysis processes a label-free quantitative (LFQ) proteomics dataset from an immunoprecipitation (IP) from ARPE19-pLVX-CLN3innHA cells targeting CLN3, a lysosomal transmembrane protein. The dataset (PXD031582) was acquired on an Orbitrap/Q-Exactive mass spectrometer and processed with MaxQuant.

Calcagni’, A., Staiano, L., Zampelli, N. et al. Loss of the batten disease protein CLN3 leads to mis-trafficking of M6PR and defective autophagic-lysosomal reformation. Nat Commun 14, 3911 (2023). https://doi.org/10.1038/s41467-023-39643-7

Data obtained from: https://www.ebi.ac.uk/pride/archive/projects/PXD031582

The experiment compares three conditions — Control (CTRL), Wild-Type CLN3 (WT), and CLN3 Knockout (KO) — with 4 biological replicates each (12 samples total), to identify proteins that interact differentially with CLN3.

## Environment Setup and Global Configuration

Clear the workspace and configure organism-specific parameters that propagate to all downstream scripts (KEGG, Gene Ontology, annotation databases). The `global_variables.R` script sets analysis thresholds (p-value, fold-change) and creates the output directory structure.

```{r setup, message=FALSE, warning=FALSE}
rm(list = ls())
gc()

# --- Organism Configuration ---
# These parameters control species-specific annotations across the entire pipeline:
#   kegg_organism: KEGG species code (dre = Danio rerio, hsa = Homo sapiens, mmu = Mus musculus)
#   species:       NCBI taxonomy ID 9606 for Human, 10090 for mouse, 7955 for zebrafish
#   organism:      Bioconductor OrgDb annotation package "org.Mm.eg.db", "org.Dr.eg.db",  "org.Hs.eg.db"
kegg_organism <- "mmu"
species <- 10090
organism <- "org.Mm.eg.db"


# --- Output Configuration ---
output_path <- "output/blood_cells_human_mouse"
image_number <- 1  # Sequential counter for figure numbering

# Load global thresholds (p_val, FC) and helper functions (create_directories)
source("../code/global_variables.R")
source("../code/aux_functions.R")
```

## Loading Pipeline Functions

Source the core pipeline modules. These scripts define the functions used throughout the analysis but do not execute any computation at this stage.

```{r load-functions, message=FALSE, warning=FALSE}
#renv::restore()
source("../code/00_packages.R")
```

## Reading Raw Data

Load peptides table. 


```{r}
peptides_data <- read_tsv("../rawdata/report_170626_PRE2026_006_SAX_15cm_AlejandraPlasma.pr_matrix.tsv")
# Clean column names (remove path prefix)
prefix <- "\\\\pcproteo02\\Data\\Raw Data\\2026\\06_Junio_26\\170626_PRE2026_006_SAX_15cm_AlejandraPlasma\\"
colnames(peptides_data) <- sub(prefix, "", colnames(peptides_data), fixed = TRUE)

# Identify sample columns (!metadata)
meta_cols <- c("Protein.Group", "Protein.Ids", "Protein.Names", "Genes",
                 "First.Protein.Description", "Proteotypic",
                 "Stripped.Sequence", "Modified.Sequence",
                 "Precursor.Charge", "Precursor.Id")
sample_cols <- setdiff(colnames(peptides_data), meta_cols)

# For every Protein.Group + sample, count unique peptides Stripped.Sequence detected (no-NA)
pep_counts <- peptides_data %>%
    select(Protein.Group, Stripped.Sequence, all_of(sample_cols)) %>%
    pivot_longer(cols = all_of(sample_cols), names_to = "sample", values_to = "intensity") %>%
    filter(!is.na(intensity)) %>%
    group_by(Protein.Group, sample) %>%
    summarise(n_peptides = n_distinct(Stripped.Sequence), .groups = "drop")

# Remain only with Protein.Groups with at least one sample with >= 2 peptides
proteins_2pep <- pep_counts %>%
    group_by(Protein.Group) %>%
    filter(any(n_peptides >= 2)) %>%
    pull(Protein.Group) %>%
    unique()

length(proteins_2pep)  # Passing filter
  
cat("====== Summary of Peptide Counts ===\n")
cat("Total number of proteins in the dataset: ", length(unique(peptides_data$Protein.Group)), "\n")
cat("Number of proteins with >= 2 peptides in at least one sample: ", length(proteins_2pep), "\n")
```


Load the MaxQuant ProteinGroups output and apply initial preprocessing: zero values are replaced with NA (MaxQuant encodes missing LFQ intensities as 0), column names are standardized across different MaxQuant versions, and known contaminant proteins are identified and removed.

```{r load-data}
prot_data <- read_tsv("../rawdata/report_170626_PRE2026_006_SAX_15cm_AlejandraPlasma.pg_matrix.tsv")

 prot_data[prot_data == 0] <- NA
colnames(prot_data)[which(names(prot_data) == "First.Protein.Description")] <- "Description"
colnames(prot_data)[which(names(prot_data) == "Protein.Group")] <- "ID"
colnames(prot_data)[which(names(prot_data) == "Genes")] <- "genename"

prefix <- "\\\\pcproteo02\\Data\\Raw Data\\2026\\06_Junio_26\\170626_PRE2026_006_SAX_15cm_AlejandraPlasma\\170626_Plasma"
  colnames(prot_data) <- sub(prefix, "", colnames(prot_data), fixed = TRUE)
  
  
prot_data <- prot_data %>%
  filter(ID %in% proteins_2pep)  # keep only proteins with >= 2 peptides in at least one sample   
```


# Experiment Design

Define the full experimental layout. This is the only section that requires manual modification when adapting the pipeline to a new dataset. All downstream scripts use this design matrix to assign samples to conditions and generate the appropriate statistical contrasts.

The design specifies:

- **Comparisons**: pairwise contrasts to test (CTRL vs WT, CTRL vs KO, KO vs WT)
- **Labels**: original MaxQuant LFQ column names used to locate intensity data
- **Conditions & Replicates**: grouping factors for the linear model

```{r experiment-design}
# --- Pairwise Comparisons ---
# Each string defines a limma contrast: "GroupA_vs_GroupB"
comparisons <- c("CTRL_vs_CTRL_K", "MDD_vs_MDD_K", "CTRL_vs_MDD", "CTRL_K_vs_MDD_K", "CTRL_vs_MDD_K")

# --- Sample Mapping ---
# label: exact column names in the MaxQuant output (used to extract LFQ intensities)
label <- c(
  #Control
"1_C_1_15cm_Slot2-1_1_5584.d",
"2_C_2_15cm_Slot2-2_1_5557.d",    
"3_C_3_15cm_Slot2-3_1_5575.d",    
"4_C_4_15cm_Slot2-4_1_5571.d",    
"5_C_5_15cm_Slot2-5_1_5553.d",    
# Control + K
"6_CK_1_15cm_Slot2-6_1_5580.d",   
"7_CK_2_15cm_Slot2-7_1_5554.d",   
"8_CK_3_15cm_Slot2-8_1_5569.d",   
"9_CK_4_15cm_Slot2-9_1_5581.d",   
"10_CK_5_15cm_Slot2-10_1_5551.d", 
# MDD
"11_MDD_1_15cm_Slot2-11_1_5579.d",
"12_MDD_2_15cm_Slot2-12_1_5558.d", 
"13_MDD_3_15cm_Slot2-13_1_5566.d", 
"14_MDD_4_15cm_Slot2-14_1_5572.d", 
"15_MDD_5_15cm_Slot2-15_1_5560.d", 
"16_MDD_6_15cm_Slot2-16_1_5559.d", 
"17_MDD_7_15cm_Slot2-17_1_5582.d", 
"18_MDD_8_15cm_Slot2-18_1_5583.d", 
# MDD+K
"19_MDDK_1_15cm_Slot2-19_1_5565.d",
"20_MDDK_2_15cm_Slot2-20_1_5552.d",
"21_MDDK_3_15cm_Slot2-21_1_5555.d",
"22_MDDK_4_15cm_Slot2-22_1_5568.d",
"23_MDDK_5_15cm_Slot2-23_1_5567.d",
"24_MDDK_6_15cm_Slot2-24_1_5574.d",
"25_MDDK_7_15cm_Slot2-25_1_5573.d",
"26_MDDK_8_15cm_Slot2-26_1_5561.d")

label_columns <- label

# columns_to_rename: short names assigned to each sample for internal use
columns_to_rename <- c(
  "CTRL_1", "CTRL_2", "CTRL_3", "CTRL_4","CTRL_5",
  "CTRL_K_1", "CTRL_K_2", "CTRL_K_3", "CTRL_K_4","CTRL_K_5",
  "MDD_1",  "MDD_2","MDD_3","MDD_4","MDD_5","MDD_6","MDD_7","MDD_8",
  "MDD_K_1",  "MDD_K_2","MDD_K_3","MDD_K_4","MDD_K_5","MDD_K_6","MDD_K_7","MDD_K_8"
)

# condition: grouping factor (must match the prefixes used in comparisons)
condition <- as.factor(c(
  "CTRL", "CTRL", "CTRL", "CTRL","CTRL",
  "CTRL_K", "CTRL_K","CTRL_K","CTRL_K","CTRL_K",
  "MDD", "MDD","MDD","MDD","MDD","MDD","MDD","MDD",
  "MDD_K", "MDD_K","MDD_K","MDD_K","MDD_K","MDD_K","MDD_K","MDD_K"
))

replicate <- c(
  "1", "2", "3", "4","5",
  "1", "2","3","4","5",
  "1",  "2","3","4","5","6","7","8",
  "1",  "2","3","4","5","6","7","8"
)

batch <- rep(1,length(replicate))

# --- Build and Export Design Matrix ---
Exp_design <- cbind.data.frame(label, condition, replicate, batch, columns_to_rename)
print(Exp_design)

Exp_design %>%
  writexl::write_xlsx(file.path(output_path, "tables","experiment_design.xlsx"))
```

## Generate SummarizedExperiment Object

```{r}
source("../code/qc_proteomics.R")

data_se <- create_se(prot_data, Exp_design)

# Save Checkpoint 
data_se |> base::saveRDS(file.path(output_path, "RData", "data_SE.RDS"))

rm(prot_data, peptides_data, pep_counts, proteins_2pep, prefix, sample_cols, meta_cols, label, columns_to_rename, condition, replicate, batch)
```


# QC

Filter, normalize, and impute missing values. This step is critical in LFQ proteomics because missing values arise from two distinct mechanisms:

- **MNAR** (Missing Not At Random): proteins below the instrument detection limit. These are imputed with low-intensity values (`zero`, `MinProb`, or `QRILC`).
- **MAR** (Missing At Random): stochastic sampling failures. These are imputed using kNN from observed values.


## Protein overlap

```{r}
plot_overlap <- plot_id_overlap(data_se)
print(plot_overlap)
save_plot("protein_overlap", plot_overlap, width = 8, height = 6, subdir = "QC")
rm(plot_overlap)
```

## Filtering Proteins with Excessive Missing Values

Fraction-based variant of filter_missing(). A protein is retained if at least one condition has a proportion of valid (non-NA) replicates >= min_valid_fraction. This threshold scales automatically with replicate count.
```{r}
data_filt <- filter_missing_frac(data_se, min_valid_fraction = 0.75)
# Save filtered NA dataset
base::saveRDS(data_filt, file.path(output_path, "RData", "data_filtered.RDS"))

# Number of proteins before filtering
cat("=== Number of protein pre-filtering = ", length(unique(rowData(data_se)$ID)),"\n")  
# Number of proteins before filtering    

cat("=== Number of protein post-filtering = ", length(unique(rowData(data_filt)$ID)),"\n")  

upset <- plot_upset_custom(data_se)
upset$table |> 
write_xlsx(file.path(output_path, "tables", "intersection_proreins.xlsx"))
upset$table |> 
  dplyr::arrange(Num_Proteins) |> 
  dplyr::slice_head(n=14)

p1 <- upset$plot
p2 <- plot_upset_custom(data_filt)

# Combined upset plot after filtering
# 
# Convert to grobs
grob1 <- grid.grabExpr(draw(p1))
grob2 <- grid.grabExpr(draw(p2$plot))

# Combine using patchwork
p <- wrap_plots(grob1, grob2, ncol = 2)+plot_annotation(
  title = "Common genes between conditions",
  subtitle = "Prefilter vs Filtered",
  tag_levels = "A",
  theme = waiver()
)
print(p)

save_plot("upset_plot", p, width = 8, height = 6, subdir = "QC")

rm(upset, p1, p2, grob1, grob2, p)
```

## Identificated proteins and coverage

```{r}
plot_sample_numbers <- plot_sample_numbers(data_filt)
print(plot_sample_numbers)

plot_protein_coverage <- plot_protein_coverage(data_filt)
print(plot_protein_coverage)

rm(plot_sample_numbers, plot_protein_coverage)
```

## Missing values distribution

```{r}
missing_heatmap_se <- plot_missing_heatmap(data_se)
missing_heatmap_filt <- plot_missing_heatmap(data_filt)
# Capture heatmap as grob from system grid
g_se   <- grid::grid.grabExpr(draw(missing_heatmap_se))
g_filt <- grid::grid.grabExpr(draw(missing_heatmap_filt))

# Mix with any ggplot
p_na <- (wrap_elements(g_se) | wrap_elements(g_filt)) + plot_annotation(
  title = "Missing values distribution",
  subtitle = "Prefilter vs Filtered",
  tag_levels = "A",
  theme = waiver()
)
save_plot("missing_values_heatmap", p_na, width = 12, height = 6, subdir = "QC")
print(p_na)

rm(missing_heatmap_se, missing_heatmap_filt, g_se, g_filt, p_na)
```

# Normalization PRONE

`PRONE::get_normalization_methods()` returns 19 identifiers. Each normalizes the assay indicated ("raw" = linear-scale intensities, "log2" = log2-transformed) as follows:

- **GlobalMean** / **GlobalMedian**: total-intensity (sample-sum) normalization; each sample is rescaled by the mean/median of the summed intensities across all samples. Input: raw.
- **Median** / **Mean**: each protein's intensity is divided by its sample's median/mean intensity, then rescaled by the mean of those medians/means across samples. Input: raw.
- **Quantile**: forces all samples to share the same intensity distribution by replacing each value with the mean of its rank-matched quantile (`preprocessCore::normalize.quantiles`). Input: log2.
- **VSN**: Variance Stabilizing Normalization (`limma::normalizeVSN`); fits a per-sample affine + generalized-log transform so variance no longer depends on intensity level. Input: raw.
- **LoessF**: single-pass ("fast") Loess normalization against the across-sample mean intensity as reference. Input: log2.
- **LoessCyc**: cyclic Loess (`limma::normalizeCyclicLoess`); MA-transforms and Loess-smooths every pair of samples, iterating over all pairs. Input: log2.
- **RLR**: Robust Linear Regression (NormalyzerDE); regresses each sample against the across-sample median and subtracts the fitted robust-linear trend. Input: log2.
- **RlrMA**: like RLR, but performed on MA-transformed data (A = median sample, M = difference from A) before the robust regression. Input: log2.
- **RlrMACyc**: cyclic variant of RlrMA — MA-transform + robust regression repeated pairwise across all sample pairs over several iterations, no fixed reference. Input: log2.
- **EigenMS**: fits an ANOVA model using the known experimental design (condition/replicate) to estimate biological effects, then applies SVD to the residuals to remove systematic non-biological trends. Input: log2.
- **MAD**: subtracts each sample's median and divides by its Median Absolute Deviation, centering and scaling all samples to a common spread. Input: log2.
- **RobNorm**: robust-regression normalization using a weighted, protein-specific density to down-weight outlier proteins when estimating per-sample scaling factors (tunable via `gamma.0`; useful for small n). Input: log2.
- **TMM**: Trimmed Mean of M-values (`edgeR::calcNormFactors`); trims extreme log-ratios/highly-abundant proteins before averaging to estimate a per-sample scaling factor (originally developed for RNA-seq). Input: raw.
- **limBE**: `limma::removeBatchEffect`; fits and subtracts a linear batch term. **Requires `metadata(se)$batch` to hold the name of a `colData` column with ≥2 batch levels.** Not applicable here — see note below.
- **IRS** (not selected below): Internal Reference Scaling; uses common bridge/reference channels replicated across TMT batches to rescale all batches onto one intensity scale. Requires both `metadata(se)$batch` and `metadata(se)$refs`; only applies to multi-batch TMT designs.
- **NormicsVSN** / **NormicsMedian**: rank proteins by a combination of mean pairwise correlation, coefficient of variation and correlation-variance to select the `top_x` most "housekeeping-like" (stable) proteins, then run VSN / median normalization using only that reference subset. Input: raw.

**Method selection for this dataset.** `limBE` and `IRS` are excluded from `methods`: `Exp_design$batch` is constant (`= 1` for all 26 samples, see the Experiment Design section), so there is no batch effect to remove, and IRS additionally requires TMT reference channels that this label-free (LFQ/DIA) dataset does not have. `PRONE::normalize_se()` would otherwise abort with `"No batch specified!"` because `metadata(se)$batch` (a column name, not a batch vector) was never set.

```{r}
PRONE::get_normalization_methods()

methods <- c("GlobalMean", "GlobalMedian", "Median",  "Mean", "Quantile",
             "VSN", "LoessF", "LoessCyc",  "RLR", "RlrMA",  "RlrMACyc",
             "EigenMS", "MAD",  "RobNorm",  "TMM",
             "NormicsVSN", "NormicsMedian")

# PRONE requires assays named "raw" and "log2" (see PRONE::load_data());
# data_filt only carries "intensity" (linear scale), so both are added here.
# rownames are stripped: PRONE's own normalization outputs are data.tables
# (which never carry rownames), and some of its internal helpers
# (e.g. get_complete_pca_dt()'s zero-variance filter) use
# identical(rm, integer(0)) to decide whether to drop rows. If our assay
# still has rownames, `which(...)` returns a *named* integer(0), which is
# no longer identical() to integer(0) — the guard misfires and the assay
# gets filtered down to 0 rows, breaking plot_PCA() with "a dimension is
# zero" for that specific assay only.
raw_mat  <- SummarizedExperiment::assay(data_filt, "intensity")
log2_mat <- log2(raw_mat)
rownames(raw_mat)  <- NULL
rownames(log2_mat) <- NULL

se_prone <- data_filt
SummarizedExperiment::assays(se_prone, withDimnames = FALSE) <- S4Vectors::SimpleList(
  raw  = raw_mat,
  log2 = log2_mat
)

# EigenMS and NormicsVSN/NormicsMedian read the protein identifier from a
# rowData column literally named "Protein.IDs" (not "ID"), and EigenMS reads
# the grouping factor from the colData column name stored in metadata$condition.
SummarizedExperiment::rowData(se_prone)$Protein.IDs <- SummarizedExperiment::rowData(data_filt)$ID
S4Vectors::metadata(se_prone)$condition <- "condition"

# The plotting functions (get_complete_dt(), used by plot_boxplots/plot_densities/...)
# melt each assay and merge it back against colData by a column literally named
# "Column" holding the sample names used as assay column names.
SummarizedExperiment::colData(se_prone)$Column <- colnames(se_prone)

se_norm <- PRONE::normalize_se(se_prone, methods = methods)

# Save RDS
se_norm |> base::saveRDS(file.path(output_path, "RData", "normalized_data.RDS"))



rm(data_se, data_filt, raw_mat, log2_mat, se_prone)
```

## QC Normalization
### Boxplots

```{r}
# Box plots
boxplots_normalization <- plot_boxplots(se_norm, ncol = 3, facet_norm = TRUE, label_by = "label", color_by = "condition") + 
  ggplot2::scale_fill_manual(values = my_colors) +
  ggplot2::theme(legend.position = "bottom", legend.direction = "horizontal")
print(boxplots_normalization)
save_plot("boxplots_normalization", boxplots_normalization, subdir = "QC", height = 27)
rm(boxplots_normalization)
```

### Density Plots

```{r}
# Density plots
density_plots <- plot_densities(se_norm, facet_norm = TRUE,  color_by = "condition") + 
  ggplot2::scale_color_manual(values = my_colors) +
  ggplot2::theme(legend.position = "bottom", legend.direction = "horizontal")
print(density_plots)
save_plot("density_plots", density_plots, subdir = "QC", height = 27)
rm(density_plots)
```

### PCA Plots

```{r}
# PCA plots
for (a in setdiff(names(SummarizedExperiment::assays(se_norm)), "raw")){
  cat("== ", a, " ==\n")
  dt <- SummarizedExperiment::assays(se_norm)[[a]]
  dt <- dt[!apply(is.na(dt),1, any),]
  cat(" after NA filter:", paste(dim(dt), collapse = " x"), "\n")
  
  rm <- which(apply(t(dt), 2, stats::var) ==0)
  cat(" n zero-variance proteins: ", length(rm), "\n")
  if (!identical(rm, integer(0))) dt <- dt[-rm,]
  cat(" after var filter:", paste(dim(dt), collapse=" x "), "\n")
  
  ok <- tryCatch({ stats::prcomp(t(dt), scale=TRUE); TRUE },
                 error = function(e){cat(" prcomp ERROR:", conditionMessage(e), "\n"); FALSE})
   cat(" prcomp ok:", ok, "\n") 
  }

  
  
  

ok_assays <- setdiff(names(SummarizedExperiment::assays(se_norm)), c("raw", "log2","EigenMS"))

pca_plots <- plot_PCA(se_norm, ok_assays, 
                      color_by = "condition",
                      facet_norm = FALSE, 
                      ellipse = TRUE) 

pca_plots <- purrr::imap(pca_plots, function(p, method){
  p + 
    ggplot2::scale_color_manual(values = my_colors) +
    ggplot2::ggtitle((paste0("PCA - ", method)))
})

print(pca_plots)
purrr::iwalk(pca_plots, function(p, method){
  save_plot(paste0("pca_normalization_", method), pca_plots[[method]], subdir = "QC")
})

rm(pca_plots, ok_assays, dt)
```



## Intragroup Variation
In PRONE, you can evaluate the intragroup variation of the normalized data by using the functions plot_intragroup_correlation(), plot_intragroup_PCV(), plot_intragroup_PMAD(), and plot_intragroup_PEV().

### Intragroup similarity measured by Pearson correlation

```{r}
pearson_plot <- plot_intragroup_correlation(se_norm, ain = NULL, "condition",
                            method = "pearson")
print(pearson_plot)
save_plot("Intragroup_Pearson", pearson_plot)

```

### Intragroup variation by calculation of the pooled coefficient of variation (PCV)

```{r}
PCV_plot <- plot_intragroup_PCV(se_norm, ain = NULL, condition = "condition", diff = FALSE)
print(PCV_plot)
save_plot("Intragroup_PCV", PCV_plot)

```


### Intragroup variation by calculation of the pooled estimate of variance (PEV)

```{r}
PEV_plot <- plot_intragroup_PEV(se_norm, ain = NULL, condition = "condition", diff = FALSE)
print(PEV_plot)
save_plot("Intragroup_PEV", PEV_plot)

```


### Intragroup variation by calculation of the pooled median absolute deviation (PMAD)

```{r}
PMAD_plot <- plot_intragroup_PMAD(se_norm, ain = NULL, condition = "condition", diff = FALSE)
print(PMAD_plot)
save_plot("Intragroup_PMAD", PMAD_plot)

```

### Composite Intragroup Variation Plot

```{r}
composite_plot <- (pearson_plot + PCV_plot) / (PEV_plot + PMAD_plot) + plot_annotation(
  title = "Intragroup Variation",
  tag_levels = "A",
  theme = waiver()
)

print(composite_plot)
save_plot("Intragroup_variation", composite_plot, subdir = "QC", width = 12, height = 10)

rm(pearson_plot, PCV_plot, PEV_plot, PMAD_plot, composite_plot)
```

## Select Normalization method

```{r}
# load RDS
se_norm <- readRDS(file.path(output_path, "RData", "normalized_data.RDS"))


data_select <- select_normalization(se_norm, method = "MAD")
```

# Imputation of missing values

## Classification of MAR/MNAR


```{r}
source("../code/imputation_single_methods.R")

mar_mnar_classification <- classify_mnar_mar(se = data_select,
                                             fraction_NA = 0.6)

mnar_distribution <- plot_mnar_distribution(mar_mnar_classification)
save_plot("mnar_distribution", mnar_distribution, subdir = "QC", width = 12, height = 8, print = TRUE)

mar_mnar_classification |> 
  writexl::write_xlsx(file.path(output_path, "tables", "MAR_MNAR_classification.xlsx"))

rm(mar_mnar_classification)
```

## Imputation

```{r}
source("../code/imputation_single_methods.R")

 data_minprob       <- impute_minprob(data_select, q = 0.01)
 data_imp_man_gauss <- impute_man(data_select, shift = 1.8, scale = 0.3)
 data_imp_knn       <- impute_knn_msn(data_select, rowmax = 0.9)
 data_imp_QRILC     <- impute_qrilc(data_select)
 data_mixed         <- impute_mixed_split(data_select,
                                         fraction_NA = 0.6,
                                         factor_SD_impute = 0.05,
                                         factor_SD_scale = 0.3,
                                         k = 10,
                                         rowmax = 0.9,
                                         colmax = 0.8,
                                         maxp = 1500,
                                         seed = 1234)
 imputed_list <- list(
   MinProb = data_minprob, man = data_imp_man_gauss,
   knn = data_imp_knn, QRILC = data_imp_QRILC, Mixed = data_mixed
 )

rm(data_minprob, data_imp_man_gauss, data_imp_knn, data_imp_QRILC, data_mixed)
```


## QC Imputation

```{r}
imputaton_distribution <- plot_imputation_distributions(data_select, imputed_list)   # from qc_proteomics.R
save_plot("imputation_distribution", imputaton_distribution, subdir = "QC", width = 12, height = 8, print = TRUE)

 imputation_sd <- plot_sd_before_after(data_select, imputed_list)            # from qc_proteomics.R
 print(imputation_sd$plot)
 save_plot("imputation_sd", imputation_sd$plot, subdir = "QC", width = 12, height = 8, print = TRUE)
 
 labels <- paste0(LETTERS[seq_along(imputed_list)], ". ", names(imputed_list))
 pca_hubert <- Map(function(se, lbl) run_pca_hubert(se, title = lbl, subtitle = NULL),
                    imputed_list, labels)                                    # from qc_proteomics.R
 plot_hubert <- patchwork::wrap_plots(lapply(pca_hubert, `[[`, "plot"), ncol = 3) +
   plot_layout(guides = "collect") +
   plot_annotation(title = "PCA-Hubert Imputation Comparison",
                   subtitle = "Robust PCA outlier detection",
                   theme = theme(plot.title = element_text(size = 16, face = "bold")),
                   caption = "Samples in red deviate from the main cluster") &
   ggplot2::theme(legend.position = "bottom", legend.direction = "horizontal")

 print(plot_hubert)
 save_plot("pca_hubert", plot_hubert, subdir = "QC", width = 12, height = 8, print = TRUE)
# Plos indivial PCA Hubert 
 for(nm in names(imputed_list)){
  cat("=== PCA-Hubert for imputation method:", nm, "===\n")
  pca_hubert[[nm]]$plot |> print()
  save_plot(paste0("pca_hubert_", nm), pca_hubert[[nm]]$plot, subdir = "QC", width = 12, height = 8, print = TRUE)
}
rm(plot_hubert, pca_hubert, imputaton_distribution, imputation_sd)
```


```{r}
pca_list <- imputed_list
pca_list$No_imputation <- data_select
pca_comparison <- plot_pca_comparison(pca_list)
save_plot("pca_comparison", pca_comparison, subdir = "QC", width = 12, height = 8, print = TRUE)

#Plot individual PCA plots
for (nm in names(pca_list)){
  cat("=== PCA for imputation method:", nm, "===\n")
  pca_list[[nm]] |> plot_pca(title = nm) |> print()
  save_plot(paste0("pca_", nm), pca_list[[nm]] |> plot_pca(title = nm), subdir = "QC", width = 12, height = 8, print = FALSE)
}
rm(pca_list, pca_comparison)
```

# Differential Expression Analysis

Perform differential expression using protein-wise linear models with empirical Bayes moderation (limma), wrapped by `DEP::analyze_dep()`. The function fits a model with the design formula `~0 + condition` (no intercept), tests the specified manual contrasts, and classifies proteins as significantly up/down-regulated based on the global thresholds (`p_val` and `FC` from `global_variables.R`).

The input is the imputed `SummarizedExperiment` object from the cleaning step. Output includes a results table with log2 fold-changes, p-values, and adjusted p-values for each comparison, exported as Excel.

                               
- BH adjusted p-value
  Controls FDR ≤ alpha exactly, assuming π₀ = 1 (all proteins are null). That assumption is never literally true, so BH is always conservative — it systematically under-reports discoveries. Its guarantee is: among the 871 proteins called significant, at most 5% are expected false positives. This is the industry standard in the proteomics field.                                                                                                            
                  
- Storey q-value
  Controls FDR ≤ alpha but estimates π₀ from the data (from the flat region of the p-value histogram near 1). Since π₀ < 1 in any real experiment, this is more honest and more powerful. The q-value of a protein means: "if I draw the significance threshold here, the expected FDR among all proteins above this threshold is q." Storey designed this specifically for high-throughput biology where many true signals exist simultaneously.                       
                  
- DEP adjusted p-value / fdrtool                                                                      
  As established: this is pathologically conservative here. It is operating on eBayes-moderated t-statistics, which violate fdrtool's distributional assumptions (the eBayes shrinkage compresses t-stats toward zero, fdrtool reads this as "most are null" and overestimates π₀). You should not use this as your primary significance criterion.



```{r differential-expression}
source("../code/de_analysis.R")
data_imp_knn <- imputed_list$knn
run_de_analysis(data_imp_knn, comparisons, output_path, adjusted_method = "BH")
```

# Visualization

Generate publication-ready plots from the differential expression results:

- **Volcano plots**: log2FC vs -log10(q-value) for each comparison, with DE proteins highlighted. For q-value adjusted
- **Heatmaps**: ComplexHeatmap clustering of significant proteins across all samples
- **PCA biplots**: sample-level dimensionality reduction to assess group separation and outliers
- **Barplots**: optional per-protein intensity barplots (disabled by default for performance)

All figures are exported in dual format (TIFF for publication, PDF for vector editing) with sequential numbering.
```{r plots, fig.width=10, fig.height=8}
print_barplots <- 0  # Set to 1 to generate individual protein barplots (slow for large datasets)

source("../code/de_plots.R")

## Sample correlation + significant-protein heatmap ####
cor_heatmap <- plot_sample_correlation(data_results)
save_plot("Correlation_matrix", cor_heatmap, subdir = "DE", width = 8, height = 8)

sig_heatmap <- plot_significant_heatmap(data_results, Exp_design)
save_plot("Heatmap_significant", sig_heatmap, subdir = "DE", width = 8, height = 10)

## PCA loadings ####
save_plot("PCA_loadings_significant", plot_pca_loadings(data_results, "significant"),
  subdir = "DE", width = 8, height = 6
)
save_plot("PCA_loadings_qval", plot_pca_loadings(data_results, "significance_qval"),
  subdir = "DE", width = 8, height = 6
)

## Volcano plots (q-value) ####
for (comp in comparisons) {
  volcano <- plot_volcano_qval(data_results, comp, p_val = p_val, fc = FC)
  save_plot(paste0("volcano_", comp), volcano, subdir = "DE", width = 8, height = 6)
}

## Fold-change correlations ####
fc_plots <- plot_fc_correlations_all(data_results, comparisons)
for (nm in names(fc_plots)) {
  save_plot(paste0("FC_correlation_", nm), fc_plots[[nm]], subdir = "DE", width = 8, height = 6)
}

## Ratio heatmaps ####
save_plot("HeatMap_ratio_padj",
  plot_ratio_heatmap(data_results, comparisons,
    filter_col = "significant", title = "Fold-change ratio (BH-adjusted significant)"
  ),
  subdir = "DE", width = 4, height = 18
)
save_plot("HeatMap_ratio_qval",
  plot_ratio_heatmap(data_results, comparisons,
    filter_col = "significance_qval", title = "Fold-change ratio (Storey q-value significant)"
  ),
  subdir = "DE", width = 4, height = 18
)

# exclude_clusters defaults to none excluded — inspect the dendrogram first,
# then re-run with the cluster numbers to drop (see de_plots.R docstring)
ratio_excluded <- plot_ratio_heatmap_excluded(data_results, comparisons, filter_col = "significance_qval", k = 6)
save_plot("HeatMap_ratio_excluded", ratio_excluded$plot, subdir = "DE", width = 4, height = 18)

## Intensity heatmaps ####
save_plot("HeatMap_intensity_padj",
  plot_intensity_heatmap(data_results, filter_col = "significant", show_rownames = FALSE),
  subdir = "DE", width = 4, height = 18
)
save_plot("HeatMap_intensity_qval",
  plot_intensity_heatmap(data_results, filter_col = "significance_qval"),
  subdir = "DE", width = 6, height = 18
)

intensity_excluded <- plot_intensity_heatmap_excluded(data_results, filter_col = "significance_qval", k = 6)
save_plot("HeatMap_intensity_excluded", intensity_excluded$plot, subdir = "DE", width = 6, height = 10)

## Barplots (optional, disabled by default — set print_barplots <- 1 above) ####
if (print_barplots) {
  adj_barplots <- plot_protein_barplots_batched(data_results, Exp_design, significant_adjusted_data$name)
  for (i in seq_along(adj_barplots)) {
    save_plot(paste0("enriched_protein_barplot_adj_", i), adj_barplots[[i]], subdir = "DE", width = 8, height = 6)
  }

  pval_barplots <- plot_protein_barplots_batched(data_results, Exp_design, significant_pval_data$name)
  for (i in seq_along(pval_barplots)) {
    save_plot(paste0("enriched_protein_barplot_pval_", i), pval_barplots[[i]], subdir = "DE", width = 8, height = 6)
  }
}
```

# Gene Ontology — Over-Representation Analysis (ORA)

<!-- Identify enriched GO terms among differentially expressed proteins using `clusterProfiler::enrichGO()`. This is an ORA approach: it tests whether DE proteins are over-represented in specific GO categories compared to the full proteome background. -->

Configuration:

- `independent_UPDOWN = 1`: analyze upregulated and downregulated proteins separately (recommended for biological interpretation; set to 0 to analyze all DE proteins together)
- `ont = "ALL"`: test all three GO ontologies simultaneously (alternatives: `"BP"` Biological Process, `"CC"` Cellular Component, `"MF"` Molecular Function)

```{r go-enrichment}
# Load data results tables
data_results <- readxl::read_xlsx(path = paste0(output_path, "tables/data_results.xlsx"))

source("../code/06_GO.R")

results_list_GO <- run_GO_enrichment(
  comparisons = comparisons,
  direction = c("UP", "DOWN"),
  independent_UPDOWN = TRUE,
  diffexpressed_type = "qval", #"qval", "pval", "padj_holm_pval"
  ont = "ALL",
  organism = organism,
  output_path = output_path,
  keyType = "UNIPROT",
  pAdjustMethod = "none")


results_df_GO <- plot_GO_results(results_list_GO = results_list_GO,
                                 direction = c("UP", "DOWN"),
                                 ont = "ALL",
                                 output_path,
                                output_dir = "enrichGO",
                                showCategory = 30,
                                top_n_terms = 15,
                                width = 8,
                                height = 12)

```

# STRING Protein Interaction Networks

Query the STRING database to retrieve known and predicted protein-protein interactions among differentially expressed proteins. Uses the `STRINGdb` package to map UniProt IDs to STRING identifiers and generate interaction networks for each comparison. This helps identify functional modules and hub proteins within the DE set.

```{r string-networks}
source("../code/07_Strings.R")
```

# Gene Set Enrichment Analysis (GSEA)

Unlike ORA (section 6), GSEA does not require a predefined list of DE proteins. Instead, it ranks **all** quantified proteins by their fold-change and tests whether members of a functional category (GO term or KEGG pathway) tend to cluster at the top or bottom of the ranked list. This approach captures coordinated moderate changes that ORA might miss.

Two analyses run in sequence:

- **gseGO**: GSEA against Gene Ontology terms
- **gseKEGG**: GSEA against KEGG metabolic and signaling pathways, with pathway maps rendered via `pathview`

```{r gsea, message=FALSE, warning=FALSE}
source("../code/08_gseGO.R")
source("../code/09_gseKEGG.R")
```

# Programmatic API Queries — STRING and PANTHER

Complement the local STRING analysis (section 7) with direct API queries via `rbioapi`. This retrieves up-to-date interaction data and functional annotations from:

- **STRING**: protein-protein interaction scores, network images, and enrichment
- **PANTHER**: protein classification, subfamily assignments, and GO-slim annotations

Unlike the STRINGdb package, `rbioapi` queries the REST APIs directly, providing access to additional endpoints and the latest database versions.

```{r rbioapi-queries, message=FALSE}
source("../code/10_RBioApi_string.R")
source("../code/11_RBioApi_panther.R")
```

# Adjusted p-value Enrichment Analysis

Re-run the functional enrichment analyses (GO ORA, gseGO, gseKEGG) using a stricter significance threshold based on adjusted p-values (`p_val_low = 0.01` from `global_variables.R`). This serves as a sensitivity analysis to assess which enriched terms remain significant under more conservative multiple testing correction.

```{r enrichment-adjusted}
source("../code/13_GO_padj.R")
source("../code/14_gseKEGG_padj.R")
source("../code/15_gseGO_adj.R")
```
# VennDiagramms

Visualize the overlap of differentially expressed proteins across comparisons using Venn diagrams. This helps identify shared and unique DE proteins between contrasts (e.g., which proteins are DE in both CTRL vs WT and CTRL vs KO, but not in KO vs WT).
```{r venn-diagrams, fig.width=8, fig.height=8}
source("../code/16_venn_diagram.R")
```



# Summary Statistics

Compute and visualize summary metrics across all comparisons: total proteins quantified, number of DE proteins (up/down) per comparison, overlap between comparisons, and effect size distributions. This section provides a high-level overview of the experiment's results.

## Statistical Summary Tables

Aggregate differential expression statistics into structured tables for reporting. Two output formats are generated:

- **Combined table** (`all_stats`): a single data frame with summary metrics across all comparisons, suitable for export
- **Per-dataset tables** (`stats_list`): maintains a list structure for experiments with multiple datasets
```{r summary-stats}
source("../code/17_summary_stats_proteomics.R")
```

```{r stats-tables}
list_results <- list(data_results)
names(list_results) <- c("IP_CLN3")

# Combined summary across all comparisons
all_stats <- apply_stats_to_list(list_results, comparisons)
print(all_stats)
writexl::write_xlsx(all_stats, paste0(output_path, "tables/all_proteomic_stats.xlsx"))

# Per-dataset summary (useful when analyzing multiple experiments)
stats_list <- apply_stats_to_list_separate(list_results, comparisons)
stats_list$IP_CLN3 %>%
  kable() %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Summary Plots

Visualize the summary statistics with multiple plot types:

- `"summary"`: overview of total and DE proteins per comparison
- `"regulation"` / `"regulation_adj"`: up vs down-regulated counts (nominal and adjusted p-values)
- `"heatmap"`: significance heatmap across comparisons
- `"comparison"`: side-by-side comparison of effect sizes
- `"all"`: generate all plot types at once

```{r stats-plots, fig.width=12, fig.height=8}
plot_proteomic_stats(all_stats, plot_type = "all")
# Individual plot types for fine-grained control
#plot_proteomic_stats(all_stats, plot_type = "summary")
#plot_proteomic_stats(all_stats, plot_type = "regulation")
#plot_proteomic_stats(all_stats, plot_type = "regulation_adj")
#plot_proteomic_stats(all_stats, plot_type = "heatmap")
#plot_proteomic_stats(all_stats, plot_type = "comparison")
```

# Session Information

```{r session-info}
sessionInfo()
```
